---
title: "Estimating broad-sense heritability with a Bayesian model"
date: "2026-06-12"
execute: 
  warning: false
  message: false
---

I previously [simulated](/posts/2025-03-08-simulate-a-phenotype/) a phenotyping dataset and estimated broad-sense heritability for a normally distributed phenotype using a [likelihood-based linear model](/posts/2025-03-17-estimate-H2/) in `lme4`.
However, I am at heart a Bayesian, so I was curious how this would work with a Bayesian model.

In this post I use the same dataset to do a very similar analysis using `brms`.
`brms` is an R package that is designed to have a very similar interface to `lme4`, but is really a front end to the much deeper Stan language for fitting Bayesian heirarchical models.
That means that if you already know how to model with `lme4` you can get a Bayesian analysis going with no effort, but you also have access to much more complicated models with the same interface.

Another strength of `brms` is that Paul Bürkner has done a superb job documenting examples of how to use it for different applications.
Have a look at the [project website](https://paulbuerkner.com/brms/index.html), and especially the (many!) [vignettes](https://paulbuerkner.com/brms/articles/index.html).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('brms')
suppressPackageStartupMessages(library(tidyverse))
```

## The phenotype data

```{r import-sim, eval=TRUE, include=FALSE}
source("https://raw.githubusercontent.com/ellisztamas/tom-ellis.info/refs/heads/main/posts/2025-03-08-simulate-a-phenotype/simulate-a-phenotype.R")
```

## Install `brms`

In principle `brms` is installed and loaded like any other R package

```r
install.packages('brms')
library('brms')
```

In practice, it can be challenging to get it to work because it relies on a huge stack of dependencies.
On my machine (running Ubuntu 24) I had to downgrade my version of R to 4.5.2 because `Rcpp` would not cooperate with R 4.6.
If you are also on Linux, check you have some obvious system dependencies:

```sh
sudo apt install build-essential libxml2-dev libcurl4-openssl-dev libssl-dev libv8-dev
```

If you're on another OS, your mileage may vary.

## Swap `lme4` for `brms`

The phenotype data are defined by genotype, replicate cohort (a blocking factor), and tray (a blocking factor nested in cohort); see [here](/posts/2025-03-08-simulate-a-phenotype/) for the details of how I generated them.
The effects genotypes, cohorts and tray were drawn from normal distributions with standard deviations of 0.5, 0.2 and 0.1 respectively.
So, if the analysis is working, we should recover estimates close to those values.

As outlined in [this post](/posts/2025-03-17-estimate-H2/), it is straightforward to estimate the variance components as random-intercept terms in `lme4`:

```{r}
library(lme4)

lmer(
  normal_phenotype ~ (1 | replicate) + (1 | tray) + (1 | line), 
  data = exp_design
)
```

This extract from the output of the model shows that the estimates are pretty close to the correct values:

```
 Groups    Name        Variance Std.Dev.
 line      (Intercept) 0.220817 0.46991 
 tray      (Intercept) 0.004996 0.07068 
 replicate (Intercept) 0.030970 0.17598 
 Residual              1.042678 1.02112 
 ```

To run the same model in `brms` we only need to swap the name of the command.

```{r}
brm_defaults <- brm(
  normal_phenotype ~ (1 | replicate) + (1 | tray) + (1 | line), 
  data = exp_design
)
```
 
There is a lot going on with the sampling and model output which is largely beyond the scope of this post.
I will focus on how well the model is estimating the variance components.

The summary of the model gives a few different estimates.

```{r}
summary(brm_defaults)
```
We are primarily interested in the standard deviation estimates (the lines starting with `sd` under `Multilevel Hyperparamters`).
For example, these lines give information about the estimates of the standard deviation between genotype means.

```sh
~line (Number of levels: 624) 
              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)     0.47      0.04     0.39     0.54 1.00     1013     2258
```
You can see that the estimate of 0.47 is very close to the true value, and the estimate from `lme4`.
The same is true for the estimate of the standard deviation between trays (0.07).
However, the estimate for standard deviation between replicate cohorts is more than double the true value.
The reason for this is that there are lots of lines and trays (624 and 39) respectively, so there are plenty of data points to estimate the variance components.
In contrast, there are only three replicate cohorts, so there is not much information in the data about the true standard deviation between cohorts.

## Using 'perfect' priors

A strength of Bayesian statistics is that it forces us to be explicit about our prior beliefs about the things we want to estimate.
In the previous model, we didn't follow that philosophy - we didn't define any explicit prior expectations about the parametes, and just used the implicit default settings.
That gave us a poor answer.
In fact, `brms` was telling us that the model was a poor fit - we got a warning about 'divergent transitions'.
In this case, it means that the default priors are not appropriate here.
That is actually quite a deep but subtle advantage of Bayesian models - when the model is poor, the model tends to warn you about that by failing, rather than letting you waste time later messing with poor results from a bad model.

In this case, we already know that the true standard deviation between replicate cohorts is 0.2.
To demonstrate that poor prior choice is what causes the divergent transitions, we can explicitly set a strong prior on the standard deviation between replicates that tells the model to look in the correct area of sampling space.
 
```{r}
# Define the prior for the standard deviation of the replicate group
priors <- c(
  set_prior("normal(0, 0.2)", class = "sd", group = "replicate")
)

# Fit the model with the defined prior
brms_perfect prior <- brm(
  normal_phenotype ~ (1 | replicate) + (1 | tray) + (1 | line), 
  data = exp_design,
  prior = priors
)
```

The summary of the model shows that the standard deviation between replicates is now estimated accurately.
Furthermore, the warning about divergent transitions has disappeared.

```{r}
summary(brms_perfect)
```
This is of course a bit silly - usually we wouldn't know the correct answer.
I just wanted to demonstrate that better priors can 'fix' a failing model.

## Priors that are too strong

Let's consider another 

```{r}
# Define the prior for the standard deviation of the replicate group
priors <- c(
  set_prior("normal(0, 0.2)", class = "sd")
)

# Fit the model with the defined prior
brms_wrong_priors <- brm(
  normal_phenotype ~ (1 | replicate) + (1 | tray) + (1 | line), 
  data = exp_design,
  prior = priors
)
```
 
## Picking priors in real life

```{r}

exp_design %>% 
  group_by(line) %>% 
  summarise(z = mean(normal_phenotype)) %>% 
  pull(z) %>% 
  sd()

exp_design %>% 
  group_by(replicate) %>% 
  summarise(z = mean(normal_phenotype)) %>% 
  pull(z) %>% 
  sd()

exp_design %>% 
  group_by(tray) %>% 
  summarise(z = mean(normal_phenotype)) %>% 
  pull(z) %>% 
  sd()
```
```{r}
# Define the prior for the standard deviation of the replicate group
priors <- c(
  set_prior("normal(0, 0.75)", class = "sd", group = "line"),
  set_prior("normal(0, 0.18)", class = "sd", group = "replicate"),
  set_prior("normal(0, 0.23)", class = "sd", group = "tray")
)

# Fit the model with the defined prior
brms_best_guess <- brm(
  normal_phenotype ~ (1 | replicate) + (1 | tray) + (1 | line), 
  data = exp_design,
  prior = priors
)
```

